Concurrent Programming with CXL

### 2020-06-20 book/paper.tex
Latest

Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
You can put the first three methods in its own CXL source file
and include it using the CXL \texttt{import} statement.
\index{import statement}
\index{module}
This would
make the code re-usable by other applications.

@@

While the CVM does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It is called \emph{spinlock},
\index{spinlock}
because each process is ``spinning'' until
it can acquire the lock.
The program uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

$I_1$ is an inductive invariant.
To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

>>>>

Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \texttt{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(\^{}pm).turn} is the value of the \texttt{.turn} key
in the mutex dictionary.  Note that the parentheses in this expression
are needed, as \texttt{\^{}pm.turn} would wrongly evaluate to
\texttt{\^{}(pm.turn)}.

You can put the first three methods in its own CXL source file
and include it using the CXL \texttt{import} statement.
\index{import statement}
\index{module}
This would
make the code re-usable by other applications.

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
atomic_inc(&count);
\end{Verbatim}
\end{code}

@@

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
The approach is called \emph{spinlock},
\index{spinlock}
because each process is ``spinning'' until
it can acquire the lock.
The program uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

$I_1$ is an inductive invariant.
To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}
(because of the condition on the \texttt{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.
